---
title: "MarketBasketAnalysis"
author: "Carlton Hencey"
date: "December 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##For this project, the purchase history at a grocery store was analyzed to determine associations between differet products. This dataset contained 9,825 different transactions, each with a collection of different products. The data was read in using the read.transactions command, and each item in the basket was separated by a comma. 
```{r Data Enter}
library(arules)
Orders<-read.transactions("groceries.csv",sep=",")
```


##The first part of this process was to determine the details of all of the purchases. First, the summary of all the purchases was obtained. Through the summary, it can be seen that the distribution of the number of items per transactionis centered mostly around values under 10, with the most frequent number of items purchased of just 1 and a median number of items purchased of 3. Also, the most frequent items purchased are provided, which shows that the most frequent item purchased is Whole Milk, followed by Other Vegetables, Rolls/Buns, Soda, and Yogurt. For further visibility on the most frequent items, a plot is included to show the top 20 most frequent items. 
```{r}
summary(Orders)
itemFrequencyPlot(Orders, topN = 20)
```

#The next step was to determine the basket rules. The interpretation of the basket rules are given through the support, confidence, and lift. The support is the percentage of transactions that contain all of the items in an itemset, the confidence is the probability that a transaction that contains the items on the left hand side of the rule also contains the item on the right hand side, and the lift is the probability of all of the items in a rule occurring together.
```{r}
Rules <- apriori(Orders, parameter = list(support =
                          0.005, confidence = 0.2, minlen = 2))
```


##The first rules that were inspected were the 5 rules with the highest support. Since the support is centered around the all items appearing together the most, it makes sense that the most frequently occuring items (whole milk, other vegetables, and rolls) also have the highest support together.
```{r}
inspect(sort(Rules, by = "support")[1:5])
```


##The next rules that were examined were the 5 rules with the highest confidence. Each of these rules contains Whole Milk on the right hand side, which indicates that each of the item groups on the left hand side has a very strong association with milk.
```{r}
inspect(sort(Rules, by = "confidence")[1:5])
```

##The top 5 rules for lift were examined next. The main takeway from these association is that root vegetables have a much higher chance of being in a basket if herbs or citrus fruit, other vegetables, and whole milk are in the basket, while whipped/sour cream has a much higher chance of being in a basket if berries or butter and other vegeatbles are in the basket. 
```{r}
inspect(sort(Rules, by = "lift")[1:5])
```

##Lastly, since Whole Milk was the most frequently purchased item, a closer look was taken at the lift of whole milk. Through the top 5 lifts involving Whole Milk, it can be seen that the baskets involving milk also have a much higher likelihood of having root vegetables, butter, and tropical fruit. 
```{r}
milkrules <- subset(Rules, items %in% "whole milk")
inspect(sort(milkrules, by = "lift")[1:5])
```

##In conclusion, these association rules can be used by the grocery store to optimize their marketing and coupon strategies. By aligning coupon strategies with lift associations, the grocery can capitalize off of customer behaviors that are already occuring. 






